\LARGE \bf Machine Translation for Tibetan Buddhist Texts

### 2021-04-07 paper/main.tex
Edited Intro and added to Disc.

Our project aims to increase the accessibility of Tibetan religious texts by training a machine translation model specifically to translate religious Tibetan texts to English. We have trained our models in PyTorch and have leveraged the BART architecture in the HuggingFrance transformers library to make the training and architecture more straightforward. By creating thesee models, we hope to motivate scholars to study not only these texts but Tibetan as a whole. Currently, while there has been some interest in machine translation for Tibetan-Chinese and vice versa, the same interest has not been reflected in Tibetan-English machine translation. From a deep dive into the open-source Tibetan NLP material, we were unable to find any projects attempting to translate Tibetan-English nor could we find papers specifically discussing Tibetan-English machine translation. By having a readily available set of translated discourses could create an on-ramp for further language uses because increased access could lead to increased interest, especially in the English-speaking community.

@@

>>>>

Digital language justice is the right for an user to use the language they are most comfortable with in the digital sphere. Language justice goes beyond this, however. With modern technology, it is possible to reconnect people with their cultural dispora. Our project aims to increase the accessibility of Tibetan religious texts by training a machine translation model specifically to translate religious Tibetan texts to English. We hope to create a translation device for Tibetan Buddhist practitioners who are non-Tibetan speakers, allowing them access to texts that had previously been barred by the language barrier. We have trained our model in PyTorch and have leveraged the BART architecture in the HuggingFrance transformers library to make the training and architecture more straightforward. By creating thesee models, we hope to motivate scholars to study not only these texts but Tibetan as a whole. Currently, while there has been some interest in machine translation for Tibetan-Chinese and vice versa, the same interest has not been reflected in Tibetan-English machine translation. From a deep dive into the open-source Tibetan NLP material, we were unable to find any projects attempting to translate Tibetan-English nor could we find papers specifically discussing Tibetan-English machine translation. By having a readily available set of translated discourses could create an on-ramp for further language uses because increased access could lead to increased interest, especially in the English-speaking community.

@@

Considering our use of word segmentation for preprocessing our data (as opposed to character segmentation), our results are comparable with, albeit lower than, Tibetan-Chinese models. Lai, Zhao, and Bao ran several different Tibetan-Chinese models on the same data using two different preprocessing methodologies, character segmentation and word segmentation. Their word segmented Transformer model averaged a higher BLEU score than ours, scoring a 38.44\% while our model scored 23.01\%.
